Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-12-21 supplement.tex
composite likelihood and Gaussian

%\sgcomment{I feel like we should improve the discussion of relationship
%with previous estimates below, or remove.}
%
%In addition to archaeological evidence, previous genetic dating suggested 
%earlier dates for a split between European and African populations. Most of 
%these results focused on populations of recent West African ancestry. 
%
%Additionally, most previous
%estimates of the time of the OOA expansion compared Eurasian and West African
%populations. Our inferred time of the East/West African divergence aligns with
%previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
%oversimplified demographic models).

\texttt{moments-LD} uses a composite likelihood approach to simultaneously fit relative
pairwise diversity and LD statistics over a range of recombination distances.
Likelihoods were computed independently for pairwise diversity and each
recombination distance bin using a multivariate Gaussian likelihood function as
described in Section \ref{sec:computing-stats}. These were multiplied across
bins and the single set of heterozygosity statistics following the approach
detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
rounds of optimization, alternating between the \emph{optimize\_log\_fmin},
\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
parameter space and hone in on the best fit parameters. Initial guesses for
parameters were chosen from demographically plausible starting points and then
perturbed to explore space, using gradient descent (on the log of the
parameters). The best fits from these initial rounds of optimization were then
chosen as the starting points for optimization using the Powell and/or the
L-BFGS-B methods (as implemented in the SciPy optimization package
\citep{Virtanen2020-kr}). This process was repeated with alternating
optimization methods until the best fit parameter set converged consistently.

\subsection{Uncertainty and confidence intervals}

To compute likelihoods, we needed an estimate of the joint
distribution of summary statistics. 
We estimated uncertainty due to the finite amount of genetic material
used in inference using bootstrap over 500 segments along the genome with
roughly equal lengths of retained sequences within each segment. First, for
each distance bin, we used these bootstrap samples to obtain a
variance-covariances matrix across all statistics. This variance-covariance
matrix was used to obtain a model likelihood for each recombination distance
bin and single-locus nucleotide diversity, as a multivariate Gaussian
likelihood. The full model likelihood was taken as the product of likelihoods
over each bin. In other words, we optimized a composite likelihood where
observations in different bins were taken to be independent.

@@

>>>>

%\sgcomment{I feel like we should improve the discussion of relationship
%with previous estimates below, or remove.}
%
%In addition to archaeological evidence, previous genetic dating suggested 
%earlier dates for a split between European and African populations. Most of 
%these results focused on populations of recent West African ancestry. 
%
%Additionally, most previous
%estimates of the time of the OOA expansion compared Eurasian and West African
%populations. Our inferred time of the East/West African divergence aligns with
%previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
%oversimplified demographic models). 
\subsection{Likelihoods, uncertainty, and confidence intervals}

\texttt{moments-LD} uses a composite multivariate Gaussian likelihood approach 
to simultaneously fit relative pairwise diversity and LD statistics over a range of 
recombination distances.
This composite likelihood and Gaussian assumption are described in more detail 
 and validated in \citet{Ragsdale2019-nt}, but we outline them here.  
To compute likelihoods, we need an estimate of the joint
distribution of summary statistics. 
We estimate uncertainty due to the finite amount of genetic material
used in inference using bootstrap over 500 segments along the genome with
roughly equal lengths of retained sequences within each segment. First, for
each distance bin, we use these bootstrap samples to obtain a
variance-covariances matrix across all statistics. This variance-covariance
matrix is used to obtain a model likelihood for each recombination distance
bin and single-locus nucleotide diversity, as a multivariate Gaussian
likelihood. 
The full model likelihood is taken as the product of likelihoods
over each bin. In other words, we followed the approach of  \citet{Ragsdale2019-nt} 
in optimizing a composite likelihood where
observations in different bins were taken to be independent.

@@

For each model tested, we ran multiple rounds of likelihood optimization, 
alternating between the \emph{optimize\_log\_fmin},
\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
parameter space and hone in on the best fit parameters. Initial guesses for
parameters were chosen from demographically plausible starting points and then
perturbed to explore space, using gradient descent (on the log of the
parameters). The best fits from these initial rounds of optimization were then
chosen as the starting points for optimization using the Powell and/or the
L-BFGS-B methods (as implemented in the SciPy optimization package
\citep{Virtanen2020-kr}). This process was repeated with alternating
optimization methods until the best fit parameter set converged consistently.
